Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-22 paper.tex
...

Pingus is an open-source puzzle-platform video game based on Lemmings.
The objective of the game is to guide a group of penguins through a number of
obstacles towards a designated exit (Figure~\ref{fig.pingus}).
% \footnote{Pingus gameplay video: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.
%
Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
of game development \cite{games.patterns}.
The codebase%
\footnote{Pingus repository: \url{https://github.com/Pingus/pingus/tree/7b255840c201d028fd6b19a2185ccf7df3a2cd6e/src}}
is about 40.000 lines of code (LoCs), divided into
the engine, level editor, auxiliary libraries, and the game logic itself.

@@

Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
to handle events from the environment (e.g., key presses and timers) and also
as a notification mechanism between entities in the game logic.
%
The observers are short-lived callbacks that have to execute as fast as
possible to keep the game reactive to incoming events in real time.
%
For this reason, callbacks cannot contain long-lasting locals and loops, which
are elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \code{goto}''.%
\footnote{``Callbacks as our Generations' Go To Statement'':
\url{http://tirania.org/blog/archive/2013/Aug-15.html}}
%\footnote{``Escape from Callback Hell'':
%\url{http://elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

@@

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting the game logic of Pingus from C++ to \CEU
with the contributions that follow:
%
\begin{itemize}
\item Identification of five recurrent control-flow patterns in the game logic
      that likely apply to other games.
\item Idiomatic solutions in \CEU for six selected behaviors in the game logic.
\item An in-depth qualitative analysis of the proposed solutions in comparison
      to C++.
\end{itemize}

\section{The Pingus Codebase}

@@

In Pingus, the game logic also accounts for almost half the size of the
codebase: 18.173 from 39.362 LoCs (46\%) spread in 272 files.
%
However, about half of the game logic relates to non-reactive code, such as
configurations and options, saved games and serialization, maps and levels
descriptions, string formatting, collision detection, graph algorithms, etc.
This part remains unchanged and relies on the seamless integration between \CEU
and C/C++~\cite{ceu.sensys13}.
%
Therefore, we rewrote 9.186 LoCs spread in 126 files%
\footnote{Game logic codebase: \url{https://github.com/an000/p/tree/master/cpp/}}.
%
In order to only consider effective code in the analysis, we then removed all
headers, declarations, trivial getters \& setters, and other innocuous
statements, resulting in 70 implementation files with 4.135 dense LoCs
originally written in C++%
\footnote{``Effective'' C++ codebase: \url{https://github.com/an000/p/tree/master/all/}}.
We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
\footnote{``Effective'' \CEU codebase: \url{https://github.com/an000/p/tree/master/all/}}.
%
Figure~\ref{tab.tree} summarizes the effective codebase in the two
implementations.

@@

\section{Control-Flow Patterns}

@@

\begin{enumerate}
\item \emph{Finite State Machines}:
    Event occurrences map to transitions between states that trigger
    appropriate actions comprising the behavior of a game entity.
\item \emph{Continuation Passing}:
    The completion of a long-lasting activity may carry a continuation, i.e.,
    some action to execute next in the game.
\item \emph{Dispatching Hierarchies}:
    Entities typically form a dispatching hierarchy in which a container that
    receives a stimulus automatically forwards it to its managed children.
\item \emph{Lifespan Hierarchies}:
    Entities typically form a lifespan hierarchy in which a terminating
    container entity automatically destroys its managed children.
\item \emph{Signaling Mechanisms}:
    Entities often need to communicate explicitly through signaling mechanisms,
    especially if there is no hierarchy relationship between them.
\end{enumerate}

@@

@@

In Pingus, a double click in the \emph{Armageddon button} at the bottom right
of the screen literally explodes all pingus.%
\footnote{Double click animation: \url{https://github.com/an000/p/blob/master/README.md#1} }
Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.

@@

The \emph{Bomber action} explodes the clicked pingu, throwing particles around
and also destroying the terrain under its radius.%
\footnote{Bomber action animation: \url{https://github.com/an000/p/blob/master/README.md#2} }
%
We can model the explosion animation with a sequential state machine
(Figure~\ref{fig.bomber}) with actions associated to specific frames as
follows%
\footnote{State machine animation: \url{https://github.com/an000/p/blob/master/README.md#3} }:
%
\begin{enumerate}
\item 0th frame:  plays a "Oh no!" sound.
\item 10th frame: plays a "Bomb!" sound.
\item 13th frame: throws particles, destroys the terrain, and shows an
                  explosion sprite.
\item Game tick:  hides the explosion sprite.
\item Last frame: kills the pingu.
\end{enumerate}

@@

The clickable \emph{blue dots} in the campaign world map transit to ambience
story screens%
\footnote{Story screen animation: \url{https://github.com/an000/p/blob/master/README.md#4} }.
A story is composed of multiple pages and, inside each page, the words of the
story appear incrementally over time.
A first click in the button \code{>>>} fast forwards the words to show the full 
page.
A second click advances to the next page, until the story terminates.
If the page completes before a click (due to the time elapsing), a first click 
advances to the next page.
%
Figure~\ref{lst.story} compares the implementations in C++ and \CEU.

@@

A pingu is a dynamic entity created periodically and destroyed under certain
conditions, such as falling from a high altitude%
\footnote{Death of pingus animation: \url{https://github.com/an000/p/blob/master/README.md#5} }.
%
Figure~\ref{lst.pingus} compares the implementations in C++ and \CEU.

@@

OptionMenu::OptionMenu() :
    conns(),
    grab_box(),
    <...>
{
    grab_box = new ChkBox(<...>);
    conns.push_back(
        cfgmgr.on_grab_change.connect(
            std::bind(&ChkBox::set, grab_box, <...>, false);
        )
    );
    conns.push_back(
        grab_box->on_change.connect(
            std::bind(&CfgMgr::set_grab, &cfgmgr, <...>);
        )
    );
    <...>

@@

OptionMenu::~OptionMenu() {
    for (Conns::iterator i=conns.begin();
                         i!=conns.end();
                         ++i) {
        (*i).disconnect();
    }
}

\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
spawn do
    var _SDL_KeyboardEvent&& e;
    every e in SDL_KEYDOWN do
        var _u8&& keys = _SDL_GetKeyState(null);
        <...>
        if e:keysym.sym == K_g then
            if keys[K_LCTRL] or keys[K_RCTRL] then
                emit cfgmgr.go_grab(
                        not {cfgmgr.get_grab()});
            end
        end
        <...>
    end
end

@@

code/await ChkBox (<...>) -> (var IChkBox chkbox) -> FOREVER do
    chkbox = val IChkBox(<...>);
    <...>
    par do
        every c.component.on_click do
            emit chkbox.go_click(not chkbox.is_on);
        end
    with
        loop do
            <...>   // switches the check box state
            chkbox.is_on = await chkbox.go_click;
        end
    end
end

@@

code/await OptionMenu <...> do
    <...>
    var& ChkBox b2 = <...>;
    spawn do
        par do
            var bool v;
            every v in cfgmgr.go_grab do
                emit b2.chkbox.go_click(v);
            end
        with
            var bool v;
            every v in b2.chkbox.go_click do
                emit cfgmgr.go_grab(v);
            end
        end
    end
    <...>
end

\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ TODO.
\label{lst.grab}
}
\end{figure*}

The \emph{mouse grab option} restricts the mouse movement to the game window
boundaries%
\footnote{Mouse grab animation: \url{https://github.com/an000/p/blob/master/README.md#6} }.
The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
In addition, the \emph{Options menu} has a check box to toggle the
\emph{mouse grab option} with mouse clicks while still responding to
\emph{Ctrl-G} presses.
%
Figure~\ref{lst.grab} compares the implementations in C++ and \CEU.

@@

In C++, the class `GlbEvt` detects \emph{Ctrl-G} presses and invokes the
callback \code{config\_manager.set\_grab} (ln. 5--8).
%
The class `CfgMgr` uses a \code{boost::signal} (ln. 16) to notify the
application when the new configuration is applied (ln. 22).
%
The \code{if} enclosing the signal emission (ln. 20--23) breaks the dependency 
cycle of Figure~\ref{fig.events} and prevents an infinite execution loop.
%
The class \code{ChkBox} also uses a \code{boost::signal} (ln. 28) to notify
the application on changes (ln. 33).
%
Again, the \code{if} enclosing the signal emission (ln. 32--34) breaks the 
dependency cycle of Figure~\ref{fig.events} to avoid infinite execution.
%
The class \code{OptionMenu} creates the dependency loop by connecting the two
signals.
%
The constructor binds
the signal \code{config\_manager.on\_grab\_change} to the callback
    method \code{grab\_box->set} (ln. 48--52),
and also
the signal \code{grab\_box->on\_change} to the callback method
    \code{config\_manager.set\_grab} (ln. 53--57).
This way, every time the \code{CfgMgr} signals
\code{on\_grab\_change} (ln. 22), the method \code{set} is implicitly called.
The same happens between the signal \code{on\_change} in the \code{ChkBox}
and the method \code{set\_grab} in the \code{CfgMgr} (ln. 18).
%
Note that the signal binding to call \code{ChkBox::set} (ln. 50) receives a
fixed value \code{false} as the last argument to prevent infinite execution
(ln. 30).
%
The \code{OptionMenu} destructor (ln. 62--66) breaks the connections explicitly
when the \emph{Option screen} terminates.

In \CEU, a \emph{Ctrl-G} key press broadcasts the internal event
\code{config\_manager.go\_grab} to the application (ln. 8).
%
The configuration manager just needs to react to \code{go\_grab}
continuously to perform the \emph{grab} effect (ln. 25--27).
%
The \code{ChkBox} exposes the event \code{go\_click} for notifications in
both directions, i.e., from the abstraction to the application and
\emph{vice versa}:
%
The abstraction reacts to external clicks continuously (ln. 41--43) to
broadcast the event \code{go\_click} (ln. 47).
It also reacts continuously to \code{go\_click} in another line of execution
(ln. 45--48), which awakes from notifications from the first line of execution
or from the application.
%
The \code{OptionMenu} connects the two events as follows:
%
The two loops in parallel handle the connections in opposite directions:
from the configuration manager to the check box (ln. 60--62);
and
from the check box to the configuration manager (ln. 65--67).
%
When the \emph{Option screen} terminates, the connections break automatically
since the body with the two loops is automatically aborted.
%
Note that the implementation in \CEU does not check event emits to break the
dependency cycle and prevent infinite execution.
Due to the stack-based execution for internal events in
\CEU~\cite{ceu.sensys13)}, programs with mutually-dependent events do not
create infinite execution loops.

\section{Related Work}

\section{Conclusion}

@@

[goto]: http://tirania.org/blog/archive/2013/Aug-15.html

>>>>

Pingus is an open-source puzzle-platform video game based on Lemmings.
The objective of the game is to guide a group of penguins through a number of
obstacles towards a designated exit (Figure~\ref{fig.pingus}).
% \footnote{Pingus gameplay video: \url{www.youtube.com/watch?v=MKrJgIFtJX0}}.
%
Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
of game development \cite{games.patterns}.
The codebase%
\footnote{Pingus repository: \url{github.com/Pingus/pingus/}}
is about 40.000 lines of code (LoCs), divided into
the engine, level editor, auxiliary libraries, and the game logic itself.

@@

Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
to handle events from the environment (e.g., key presses and timers) and also
as a notification mechanism between entities in the game logic.
%
The observers are short-lived callbacks that have to execute as fast as
possible to keep the game reactive to incoming events in real time.
%
For this reason, callbacks cannot contain long-lasting locals and loops, which
are elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \code{goto}''.%
\footnote{``Callbacks as our Generations' Go To Statement'':
\url{tirania.org/blog/archive/2013/Aug-15.html}}
%\footnote{``Escape from Callback Hell'':
%\url{elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

@@

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to \CEU with the
contributions as follows:
%
\begin{itemize}
\item Identification of five recurrent control-flow patterns that likely apply
      to other games.
\item Idiomatic code in \CEU for six selected behaviors in the game logic.
\item An in-depth qualitative analysis of the implementations in \CEU in
      comparison to C++.
\end{itemize}

TODO: falar mais sobre as 3 contribuicoes:
- control-flow patterns
    - o que eh isso
        - definir controle
    - como eh diferente de GoF
        - behavioral patterns
- ideia geral das solucoes
    - eliminar estados explicitos com estruturas de controle apropriadas
- o que eh analise qualitativa
    - como medimos

The rest of the paper is organized as follows:
Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
\CEU.
Section~\ref{sec.patterns} describes the five control-flow patterns and
discusses six case studies.
Section~\ref{sec.related} discusses related work.
Section~\ref{sec.conclusion} concludes the paper.

\section{The Pingus Codebase}
\label{sec.codebase}

@@

In Pingus, the game logic also accounts for almost half the size of the
codebase: 18.173 from 39.362 LoCs (46\%) spread across 272 files.
%
However, about half of the game logic relates to non-reactive code, such as
configurations and options, saved games and serialization, maps and levels
descriptions, string formatting, collision detection, graph algorithms, etc.
This part remains unchanged and relies on the seamless integration between \CEU
and C/C++~\cite{ceu.sensys13}.
%
Therefore, we rewrote 9.186 LoCs spread across 126 files%
\footnote{Complete codebase: \url{github.com/an000/p/tree/master/cpp}}.
%
In order to only consider effective code in the analysis, we then removed all
headers, declarations, trivial getters \& setters, and other innocuous
statements, resulting 4.135 dense LoCs spread across 70 implementation files
originally written in C++%
\footnote{C++ codebase: \url{github.com/an000/p/tree/master/all}}.
We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
\footnote{\CEU codebase: \url{github.com/an000/p/tree/master/all}}.
%
Figure~\ref{tab.tree} summarizes the effective codebase in the two
implementations.

@@

\section{Control-Flow Patterns \& Case Studies}
\label{sec.patterns}

TODO: state pattern, FSM
      strategy/visitor pattern, polymorhpism, hierarchies
      observer pattern, signaling

@@

\begin{enumerate}
\item \emph{Finite State Machines}:
    Event occurrences map to transitions between states that trigger
    appropriate actions comprising the behavior of a game entity.
\item \emph{Continuation Passing}:
    The completion of a long-lasting activity may carry a continuation, i.e.,
    some action to execute next in the game.
\item \emph{Dispatching Hierarchies}:
    Entities form a dispatching hierarchy in which a container that receives a
    stimulus automatically forwards it to its managed children.
\item \emph{Lifespan Hierarchies}:
    Entities form a lifespan hierarchy in which a terminating container entity
    automatically destroys its managed children.
\item \emph{Signaling Mechanisms}:
    Entities often need to communicate explicitly through signaling mechanisms,
    especially if there is no hierarchy relationship between them.
\end{enumerate}

@@

TODO: falar do state pattern, citar game.patterns, falar que usa estado explicito

@@

In Pingus, a double click in the \emph{Armageddon button} at the bottom right
of the screen literally explodes all pingus.%
\footnote{Double click animation: \url{github.com/an000/p/blob/master/README.md#1} }
Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.

@@

The \emph{Bomber action} explodes the clicked pingu, throwing particles around
and also destroying the terrain under its radius.%
\footnote{Bomber action animation: \url{github.com/an000/p/blob/master/README.md#2} }
%
We can model the explosion animation with a sequential state machine
(Figure~\ref{fig.bomber}) with actions associated to specific frames as
follows%
\footnote{State machine animation: \url{github.com/an000/p/blob/master/README.md#3} }:
%
\begin{enumerate}
\item 0th frame:  plays a "Oh no!" sound.
\item 10th frame: plays a "Bomb!" sound.
\item 13th frame: throws particles, destroys the terrain, and shows an
                  explosion sprite.
\item Game tick:  hides the explosion sprite.
\item Last frame: kills the pingu.
\end{enumerate}

@@

The clickable \emph{blue dots} in the campaign world map transit to ambience
story screens%
\footnote{Story screen animation: \url{github.com/an000/p/blob/master/README.md#4} }.
A story is composed of multiple pages and, inside each page, the words of the
story appear incrementally over time.
A first click in the button \code{>>>} fast forwards the words to show the full 
page.
A second click advances to the next page, until the story terminates.
If the page completes before a click (due to the time elapsing), a first click 
advances to the next page.
%
Figure~\ref{lst.story} compares the implementations in C++ and \CEU.

@@

A pingu is a dynamic entity created periodically and destroyed under certain
conditions, such as falling from a high altitude%
\footnote{Death of pingus animation: \url{github.com/an000/p/blob/master/README.md#5} }.
%
Figure~\ref{lst.pingus} compares the implementations in C++ and \CEU.

@@

OptMenu::OptMenu() :
  conns(),
  grab_box(),
  <...>
{
  grab_box = new ChkBox(<...>);
  conns.push_back(
    cfgmgr.on_grab_change.connect( std::bind(
      &ChkBox::set, grab_box, <...>, false) );
    )
  );
  conns.push_back(
    grab_box->on_change.connect( std::bind(
      &CfgMgr::set_grab, &cfgmgr, <...>) );
    )
  );
  <...>

@@

OptMenu::~OptMenu() {
  for (Conns::iterator i=conns.begin();
             i!=conns.end();
             ++i) {
    (*i).disconnect();
  }
}

.
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
spawn do
  var _SDL_KeyboardEvent&& e;
  every e in SDL_KEYDOWN do
    var _u8&& keys = _SDL_GetKeyState(null);
    <...>
    if e:keysym.sym == K_g then
      if keys[K_LCTRL] or keys[K_RCTRL] then
        emit cfgmgr.go_grab(
            not {cfgmgr.get_grab()});
      end
    end
    <...>
  end
end

@@

code/await ChkBox (<...>) -> (var IChkBox box) -> FOREVER do
  box = val IChkBox(<...>);
  <...>
  par do
    every c.component.on_click do
      emit box.go_click(not box.is_on);
    end
  with
    loop do
      <...>   // switches the check box state
      box.is_on = await box.go_click;
    end
  end
end

@@

code/await OptMenu <...> do
  <...>
  var& ChkBox b2 = <...>;
  spawn do
    par do
      var bool v;
      every v in cfgmgr.go_grab do
        emit b2.box.go_click(v);
      end
    with
      var bool v;
      every v in b2.box.go_click do
        emit cfgmgr.go_grab(v);
      end
    end
  end
  <...>
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ TODO.
\label{lst.grab}
}
\end{figure*}

The \emph{mouse grab option} restricts the mouse movement to the game window
boundaries%
\footnote{Mouse grab animation: \url{github.com/an000/p/blob/master/README.md#6} }.
The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
In addition, the \emph{Options menu} has a check box to toggle the
\emph{mouse grab option} with mouse clicks while still responding to
\emph{Ctrl-G} presses.
%
Figure~\ref{lst.grab} compares the implementations in C++ and \CEU.

@@

In C++, the class `GlbEvt` detects \emph{Ctrl-G} presses and invokes the
callback \code{config\_manager.set\_grab} (ln. 5--8).
%
The class `CfgMgr` uses a \code{boost::signal} (ln. 16) to notify the
application when the new configuration is applied (ln. 22).
%
The \code{if} enclosing the signal emission (ln. 20--23) breaks the dependency 
cycle of Figure~\ref{fig.events} and prevents an infinite execution loop.
%
The class \code{ChkBox} also uses a \code{boost::signal} (ln. 28) to notify
the application on changes (ln. 33).
%
Again, the \code{if} enclosing the signal emission (ln. 32--34) breaks the 
dependency cycle of Figure~\ref{fig.events} to avoid infinite execution.
%
The class \code{OptMenu} creates the dependency loop by connecting the two
signals.
%
The constructor binds
the signal \code{config\_manager.on\_grab\_change} to the callback
    method \code{grab\_box->set} (ln. 48--52),
and also
the signal \code{grab\_box->on\_change} to the callback method
    \code{config\_manager.set\_grab} (ln. 53--57).
This way, every time the \code{CfgMgr} signals
\code{on\_grab\_change} (ln. 22), the method \code{set} is implicitly called.
The same happens between the signal \code{on\_change} in the \code{ChkBox}
and the method \code{set\_grab} in the \code{CfgMgr} (ln. 18).
%
Note that the signal binding to call \code{ChkBox::set} (ln. 50) receives a
fixed value \code{false} as the last argument to prevent infinite execution
(ln. 30).
%
The \code{OptMenu} destructor (ln. 62--66) breaks the connections explicitly
when the \emph{Option screen} terminates.

In \CEU, a \emph{Ctrl-G} key press broadcasts the internal event
\code{config\_manager.go\_grab} to the application (ln. 8).
%
The configuration manager just needs to react to \code{go\_grab}
continuously to perform the \emph{grab} effect (ln. 25--27).
%
The \code{ChkBox} exposes the event \code{go\_click} for notifications in
both directions, i.e., from the abstraction to the application and
\emph{vice versa}:
%
The abstraction reacts to external clicks continuously (ln. 41--43) to
broadcast the event \code{go\_click} (ln. 47).
It also reacts continuously to \code{go\_click} in another line of execution
(ln. 45--48), which awakes from notifications from the first line of execution
or from the application.
%
The \code{OptMenu} connects the two events as follows:
%
The two loops in parallel handle the connections in opposite directions:
from the configuration manager to the check box (ln. 60--62);
and
from the check box to the configuration manager (ln. 65--67).
%
When the \emph{Option screen} terminates, the connections break automatically
since the body with the two loops is automatically aborted.
%
Note that the implementation in \CEU does not check event emits to break the
dependency cycle and prevent infinite execution.
Due to the stack-based execution for internal events in
\CEU~\cite{ceu.sensys13)}, programs with mutually-dependent events do not
create infinite execution loops.

\section{Related Work}
\label{sec.related}

Some previous work~\cite{a,b,c} discuss how the GoF patterns~\cite{gof} apply
to video game development.
    - generally?
        - mostram exemplos abstratos
            - imagine um jogo...
                - assim poderiamos aplicar...
            - sem codigo real
            - nao aplicam a jogos existentes

- patterns
    - sbgames
    - gpp   
- contol-flow patterns that are recurrent in Pingus and likely apply to other
games
- also, not identifying and applying, but applying in idiomatic code in another
  language

Design patterns for games can be approached from two dif-
ferent perspectives; firstly as patterns used for describing
the game mechanics (gameplay and game rules) and sec-
ondly as the use of object-oriented design patterns in pro-
gramming games.
Concerning game mechanics, Bjork et al. [4] introduced
a set of design patterns which essentially are descriptions
(employing a unified vocabulary) of reoccurring interac-
tion schemes relevant to game’s story and gameplay. As
such, these patterns are not related to the software archi-
tecture or code. The proposed patterns are collected from
interviewing professional game programmers, from ana-
lyzing existing games and from transforming game
mechanics.

\section{Conclusion}
\label{sec.conclusion}

@@

[goto]: tirania.org/blog/archive/2013/Aug-15.html
